---
title: "Report on MovieLens"
author: "Raphael Kummer"
date: "`r format(Sys.Date())`"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, progress = TRUE, verbose = TRUE)
```

# MovieLens

## Introduction

Build a ML model to recommend movies to users.

### Dataset

Grouplens created a movie rating dataset. The 10M dataset [@harper2015] used in this project is a subset of 10 million ratings of 10'000 movies by 72'000 random selected users.

## Problem

Given is the loading of the MovieLens 10M dataset, split into an *edx* and a *final_holdout_test* set containing 10% of the MovieLens data. The dataset contains userId, movieId, rating, timestamp, title, and genre.

```{r echo = TRUE, warning = FALSE}
##########################################################
# Create edx and final_holdout_test sets 
##########################################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

options(timeout = 120)

dl <- "ml-10M100K.zip"
if(!file.exists(dl))
  download.file("https://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings_file <- "ml-10M100K/ratings.dat"
if(!file.exists(ratings_file))
  unzip(dl, ratings_file)

movies_file <- "ml-10M100K/movies.dat"
if(!file.exists(movies_file))
  unzip(dl, movies_file)

ratings <- as.data.frame(str_split(read_lines(ratings_file), fixed("::"), simplify = TRUE),
                         stringsAsFactors = FALSE)
colnames(ratings) <- c("userId", "movieId", "rating", "timestamp")
ratings <- ratings %>%
  mutate(userId = as.integer(userId),
         movieId = as.integer(movieId),
         rating = as.numeric(rating),
         timestamp = as.integer(timestamp))

movies <- as.data.frame(str_split(read_lines(movies_file), fixed("::"), simplify = TRUE),
                        stringsAsFactors = FALSE)
colnames(movies) <- c("movieId", "title", "genres")
movies <- movies %>%
  mutate(movieId = as.integer(movieId))

movielens <- left_join(ratings, movies, by = "movieId")

# Final hold-out test set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.6 or later
# set.seed(1) # if using R 3.5 or earlier
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in final hold-out test set are also in edx set
final_holdout_test <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from final hold-out test set back into edx set
removed <- anti_join(temp, final_holdout_test)
```

## Data Inspection and preprocessing
```{r echo = FALSE}
install.packages("kable")
```

The *edx* dataset look like this:
```{r  echo = FALSE}
knitr::kable(head(edx))
```
```{r echo = FALSE}
head(edx)
```





## Model

## Results

### RMSE

## Conclusion

## Resources

[1] Rafael Irizarry. 2018. Introduction to Data Science. <https://rafalab.dfci.harvard.edu/dsbook/>
